A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-19 tex/codatadp.tex
more

\paragraph{Phase 3: Branching.}
If a selector is applied to a term $t$, or if $t$'s type is a finite datatype,
the equivalence class of $t$ must contain an application of a $\delta$ constructor
(where $\Ctr^\delta = \{ \const{C}_1, \ldots, \const{C}_m \}$).
This is enforced in the third phase by the \rn{Split} rule.
A companion rule, \rn{Singleton}, handles the degenerate case where two terms
from $\tEc$ have a corecursive singleton type, in which case they must be
equivalent.

\paragraph{Correctness.}
An application of a rule is \emph{redundant} if at least one branch in the
conclusion does not add a new equality to $\Ec$ and is not $\bot$.
A \emph{derivation tree} is a tree whose nodes are sets of equalities, where non-root nodes are obtained by
a nonredundant application of a derivation rule to its parent node.
A derivation tree is \emph{closed} if all of its leaf nodes are $\bot$.
A node is \emph{saturated} if no nonredundant application of a rule can be applied to it.
If there exists a closed derivation tree with root node $\Ec$, then we conclude that $\Ec$ is $\thD$-unsatisfiable.
If there exists a derivation tree with root node $\Ec$ that contains a saturated node, then we conclude that $\Ec$ is $\thD$-satisfiable.

\begin{theorem}[Termination]%
\label{thm:t}%
\afterDot
All derivation trees are finite.
\end{theorem}
\begin{proof}
Consider a derivation tree with root node $\Ec$.
Let $S$ be the set of terms occurring as the argument of a selector in $\tEc$.
Let $T$ be the set of terms in $\tEc$ of finite datatype type.
For each term $t \in T$,
let $S^0_t$ be the set $\{ t \}$,
for $i \geq 0$ let $S^{i+1}_t =$ $S^i_t \cup \{ s( u ) \mid u : \upsilon \in S^i, \upsilon \in \Data, \upsilon \text{ finite}, s \in \Sel^{\upsilon}  \}$,
and let $S^\ast_t$ be the fixed point of this sequence.
We know this is a finite set for each $t$ since the values of the type of $t$ are of finite size.
Let $S^\ast$ be the union of $S^\ast_t$ for all $t \in T$,
and let $\tcEc$ be the set of subterms of $\Ec \cup \{ C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) ) \mid t : \tau \in S \cup S^\ast, C_i \in \Ctr^\tau \}$.
In a derivation tree with root node $\Ec$,
it can be shown by case analysis on the rules of the calculus that each non-root node $\Fc$ is such that
$\mathcal{T}(\Fc) \subseteq \tcEc$, and moreover thus contains an equality between two terms from $\tcEc$ not occurring in its parent node.
Thus, the depth of a branch in a derivation tree with root node $\Ec$ is at most $\mid \tcEc \mid^2$,
which is finite since $\tcEc$ is finite.
\qed
\end{proof}

\begin{theorem}[Refutation Soundness]%
\label{thm:rs}%
\afterDot
If there exists a closed derivation tree with root node\/ $\Ec$, then\/ $\Ec$ is $\thD$-unsatisfiable.
\end{theorem}
\begin{proof}
We prove this by structural induction on the (closed) derivation tree with root node $\Ec$.
First, if the derivation tree is an application of \rn{Conflict}, \rn{Clash}, or \rn{Acyclic},
then $\Ec$ is $\thD$-unsatisfiable.
For \rn{Conflict}, this is a consequence of equality reasoning.
For \rn{Clash}, this is a consequence of distinctness.
For \rn{Acyclic}, our construction of $\Val$ indicates that the class of values that $t_1$ can take in models of $\Ec$ is infeasible,
and thus $\Ec$ is unsatisfiable.
If the children of $\Ec$ are closed derivation trees
whose root node is the result of applying the rule \rn{Split} on term $t$ of type $\tau$,
then by the induction hypothesis $\Ec \cup t \teq C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) )$ is unsatisfiable
for each $C_i \in \Ctr^\tau$.
Since by exhaustiveness, all models of $\thD$ entail (exactly) one equality $t \teq C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) )$,
we know $\Ec$ is $\thD$-unsatisfiable.
Otherwise, the child of $\Ec$ is a closed derivation tree
whose root node is $\Ec \cup t_1 \teq t_2$ obtained by applying one of the rules \rn{Refl}, \rn{Sym}, \rn{Trans}, \rn{Cong}, \rn{Inject}, \rn{Unique}, or \rn{Singleton}.
In all cases, we have that $\Ec \models_{\thD} t_1 \teq t_2$.
For \rn{Refl}, \rn{Sym}, \rn{Trans}, \rn{Cong}, this is a consequence of equality reasoning.
For \rn{Inject}, this is a consequence of injectivity.
For \rn{Unique}, our construction of $\Val$ indicates that the value of $t_1$ and $t_2$ are equivalent in all models of $\Ec$.
For \rn{Singleton}, clearly $t_1$ and $t_2$ must have the same value since the cardinality of their type is one.
By the induction hypothesis, we have $\Ec \cup t_1 \teq t_2$ is $\thD$-unsatisfiable,
and thus $\Ec$ is $\thD$-unsatisfiable.
\qed
\end{proof}

@@

For a $\mu$-term $t_1$ with subterm $t_2$,
the \emph{interpretation of $t_2$ in $t_1$} is the $\mu$-term $t_1\interp{t_2}^\emptyset$ as returned by the following recursive procedure,
where $x \mapsto \MU x. \const{C}( \vec u ) \in t_1$ indicates that $\MU x. \const{C}( \vec u )$ is the subterm of $t_1$ that binds this occurrence of variable $x$.
\[\begin{array}{r@{}c@{}l}
t_1\interp{x}^M & {} \,=\, {} &
\textstyle\begin{cases}
    \MU x.\; \const{C}( t_1\interp{ \vec u}^{M \cup \{ x \}}) & \text{if }x \not\in M, x \mapsto \MU x. \const{C}( \vec u ) \in t_1  \\[-\jot]
    x & \text{otherwise}
\end{cases} \\
t_1\interp{\MU x. \const{C}( \vec u )}^M & = &
\textstyle\begin{cases}
    \MU x. \const{C}( t_1\interp{ \vec u }^{M \cup \{ x \}} ) & \text{if }x \not\in M \\[-\jot]
    x & \text{otherwise}
\end{cases}
\end{array}\]
We often write $t_1\interp{t_2}$ as shorthand for $t_1\interp{t_2}^\emptyset$.

@@

\paragraph{Integration into DPLL(T)}
The calculus is implemented as \emph{theory solver} of CVC4,
that is, a specialized solver for determining the satisfiability of conjunctions of constraints for its theory.
Given a theory $T = T_1 \cup \ldots T_n$ and a input set of clauses $F$ in CNF,
the DPLL(T) \rem{CDCL(T)?} procedure
(incrementally) builds partial assignments from the atomic formulas of $F$ to truth values such that no clause in $F$ is falsified.
We may interpret this partial assignment as a set of literals where $a \in M$ for all atoms $a$ that it assigned the value $\top$, and $\neg a \in M$ for all atoms that it assigns the value $\bot$.
Then, at a high level, by the Nelson--Oppen combination framework,
each $T_i$-solver for $i = 1, \ldots, n$ takes as input a combination of (1) the purified form of $T_i$-constraints occurring in $M$ where terms containing symbols not belonging to $T_i$ are replaced by fresh variables of the same type,
(2) additional equalities and disequalities between variables of types not belonging to $T_i$.
Let us call this set $M_i$.
Each $T_i$ solver either
reports that a subset $C$ of $M_i$ is $T_i$-unsatisfiable, in which case $\neg C$ is added as a additional clause to $F$,
adds an additional clause $C$ to $F$,
or does nothing.
When $M$ is a complete assignment for $F$, then a theory solver can only choose to do nothing when $M_i$ is indeed $T_i$-satisfiable.

>>>>

\paragraph{Phase 3: Branching.}
If a selector is applied to a term $t$, or if $t$'s type is a finite datatype,
the equivalence class of $t$ must contain an application of one of
$\delta$'s constructors
$\Ctr^\delta = \{ \const{C}_1, \ldots, \const{C}_m \}$.
This is enforced in the third phase by the \rn{Split} rule.
A companion rule, \rn{Singleton}, handles the degenerate case where two terms
from $\tEc$ have a corecursive singleton type, in which case they must be
equivalent.

\paragraph{Correctness.}
An application of a rule is \emph{redundant} if at least one branch in the
conclusion does not add a new equality to $\Ec$ and is not $\bot$.
A \emph{derivation tree} is a tree whose nodes are sets of equalities, where non-root nodes are obtained by
a nonredundant application of a derivation rule to its parent node.
A derivation tree is \emph{closed} if all of its leaf nodes are $\bot$.
A node is \emph{saturated} if no nonredundant instance of a rule can be applied to it.
%
If there exists a closed derivation tree with root node $\Ec$, we conclude that $\Ec$ is $\thD$-unsatisfiable.
If there exists a derivation tree with root node $\Ec$ that contains a saturated node, we conclude that $\Ec$ is $\thD$-satisfiable.

\begin{theorem}[Termination]%
\label{thm:t}%
\afterDot
All derivation trees are finite.
\end{theorem}
\begin{proof}
Consider a derivation tree with root node $\Ec$.
Let $S$ be the set of terms occurring as the argument of a selector in $\tEc$.
Let $T$ be the set of terms in $\tEc$ of finite datatype type.
For each term $t \in T$,
let $S^0_t$ be the set $\{ t \}$,
for $i \geq 0$ let $S^{i+1}_t =$ $S^i_t \mathrel\cup \{ s( u ) \mid u : \upsilon \in S^i, \upsilon \in \Data, \upsilon \text{ finite}, s \in \Sel^{\upsilon}  \}$,
and let $S^\ast_t$ be the fixed point of this sequence.
We know this is a finite set for each $t$ since the values of the type of $t$ are of finite size.
Let $S^\ast$ be the union of $S^\ast_t$ for all $t \in T$,
and let $\tcEc$ be the set of subterms of $\Ec \mathrel\cup \{ C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) ) \mid t : \tau \in S \mathrel\cup S^\ast, C_i \in \Ctr^\tau \}$.
In a derivation tree with root node $\Ec$,
it can be shown by case analysis on the rules of the calculus that each non-root node $\Fc$ is such that
$\mathcal{T}(\Fc) \subseteq \tcEc$, and moreover thus contains an equality between two terms from $\tcEc$ not occurring in its parent node.
Thus, the depth of a branch in a derivation tree with root node $\Ec$ is at most $\left| \tcEc \right|^2$,
which is finite since $\tcEc$ is finite.
\qed
\end{proof}

\begin{theorem}[Refutation Soundness]%
\label{thm:rs}%
\afterDot
If there exists a closed derivation tree with root node\/ $\Ec$, then\/ $\Ec$ is $\thD$-unsatisfiable.
\end{theorem}
\begin{proof}
We prove this by structural induction on the (closed) derivation tree with root node $\Ec$.
First, if the derivation tree is an application of \rn{Conflict}, \rn{Clash}, or \rn{Acyclic},
then $\Ec$ is $\thD$-unsatisfiable.
For \rn{Conflict}, this is a consequence of equality reasoning.
For \rn{Clash}, this is a consequence of distinctness.
For \rn{Acyclic}, our construction of $\Val$ indicates that the class of values that $t_1$ can take in models of $\Ec$ is infeasible,
and thus $\Ec$ is unsatisfiable.
If the children of $\Ec$ are closed derivation trees
whose root node is the result of applying the rule \rn{Split} on term $t$ of type $\tau$,
then by the induction hypothesis $\Ec \mathrel\cup t \teq C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) )$ is unsatisfiable
for each $C_i \in \Ctr^\tau$.
Since by exhaustiveness, all models of $\thD$ entail (exactly) one equality $t \teq C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) )$,
we know $\Ec$ is $\thD$-unsatisfiable.
Otherwise, the child of $\Ec$ is a closed derivation tree
whose root node is $\Ec \mathrel\cup t_1 \teq t_2$ obtained by applying one of the rules \rn{Refl}, \rn{Sym}, \rn{Trans}, \rn{Cong}, \rn{Inject}, \rn{Unique}, or \rn{Singleton}.
In all cases, we have that $\Ec \models_{\thD} t_1 \teq t_2$.
For \rn{Refl}, \rn{Sym}, \rn{Trans}, \rn{Cong}, this is a consequence of equality reasoning.
For \rn{Inject}, this is a consequence of injectivity.
For \rn{Unique}, our construction of $\Val$ indicates that the value of $t_1$ and $t_2$ are equivalent in all models of $\Ec$.
For \rn{Singleton}, clearly $t_1$ and $t_2$ must have the same value since the cardinality of their type is one.
By the induction hypothesis, we have $\Ec \mathrel\cup t_1 \teq t_2$ is $\thD$-unsatisfiable,
and thus $\Ec$ is $\thD$-unsatisfiable.
\qed
\end{proof}

@@

For a $\mu$-term $t_1$ with subterm $t_2$,
the \emph{interpretation of $t_2$ in $t_1$} is the $\mu$-term $t_1\interp{t_2}^\emptyset$ as returned by the following recursive procedure,
where $x \mapsto \MU x. \const{C}( \vec u ) \in t_1$ indicates that $\MU x. \const{C}( \vec u )$ is the subterm of $t_1$ that binds this occurrence of variable $x$.
\[\begin{array}{r@{}c@{}l}
t_1\interp{x}^M & {} \,=\, {} &
\textstyle\begin{cases}
    \MU x.\; \const{C}( t_1\interp{ \vec u}^{M \mathrel\cup \{ x \}}) & \text{if }x \not\in M, x \mapsto \MU x. \const{C}( \vec u ) \in t_1  \\[-\jot]
    x & \text{otherwise}
\end{cases} \\
t_1\interp{\MU x. \const{C}( \vec u )}^M & = &
\textstyle\begin{cases}
    \MU x. \const{C}( t_1\interp{ \vec u }^{M \mathrel\cup \{ x \}} ) & \text{if }x \not\in M \\[-\jot]
    x & \text{otherwise}
\end{cases}
\end{array}\]
We often write $t_1\interp{t_2}$ as shorthand for $t_1\interp{t_2}^\emptyset$.

@@

\paragraph{Integration into DPLL(T)}
The calculus is implemented as \emph{theory solver} of CVC4,
that is, a specialized solver for determining the satisfiability of conjunctions of constraints for its theory.
Given a theory $T = T_1 \mathrel\cup \ldots T_n$ and a input set of clauses $F$ in CNF,
the DPLL(T) \rem{CDCL(T)?} procedure
(incrementally) builds partial assignments from the atomic formulas of $F$ to truth values such that no clause in $F$ is falsified.
We may interpret this partial assignment as a set of literals where $a \in M$ for all atoms $a$ that it assigned the value $\top$, and $\neg a \in M$ for all atoms that it assigns the value $\bot$.
Then, at a high level, by the Nelson--Oppen combination framework,
each $T_i$-solver for $i = 1, \ldots, n$ takes as input a combination of (1) the purified form of $T_i$-constraints occurring in $M$ where terms containing symbols not belonging to $T_i$ are replaced by fresh variables of the same type,
(2) additional equalities and disequalities between variables of types not belonging to $T_i$.
Let us call this set $M_i$.
Each $T_i$ solver either
reports that a subset $C$ of $M_i$ is $T_i$-unsatisfiable, in which case $\neg C$ is added as a additional clause to $F$,
adds an additional clause $C$ to $F$,
or does nothing.
When $M$ is a complete assignment for $F$, then a theory solver can only choose to do nothing when $M_i$ is indeed $T_i$-satisfiable.
